\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{hyperref}
\usepackage{cite}

\graphicspath{{figures/}}

\title{DualBalance Districting:\\From Detection to Generation}
\author{Steven Hart}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Whether it is possible to draw congressional districts without
traditional human bias is both a practical and a philosophical
question. In most states the same legislators who benefit from
district boundaries draw them; federal courts cannot review
partisan-gerrymandering claims after \textit{Rucho v.~Common Cause};
and the extensive toolkit of forensic metrics can measure manipulation
after the fact but cannot prevent it. This paper investigates one
approach: generate the map from a fully public, pre-committed rule
that eliminates within-state discretion entirely. \emph{DualBalance
Districting} is a deterministic algorithm that computes congressional
districts as a pure function of state geometry, census population,
and seat count~$N$, with no discretionary parameters and no political
inputs. Every design choice is pre-committed before any state's data
are examined. Each district carries roughly $1/N$ of the state's
people \emph{and} $1/N$ of its land, weighted equally in the
DualBalance Score (DBS). Because no within-state line-drawer makes
any choices, the conventional forensic metrics (Efficiency Gap, shape
compactness, ensemble outlier tests) retain their descriptive validity
on DualBalance maps but lose their inferential role as evidence of
intent. The inference from shape to motive requires a line-drawer;
this construction has none.

Validated on all 41 multi-seat states with TIGER 2020PL VTD data,
DualBalance achieves \emph{Karcher}-compliant population balance
($\mathrm{pop\_dev\_max} \leq 0.05\,\%$) on 26 states; three states
(Florida, New York, West Virginia) are known geometric failures
requiring multi-center seeding as a structural fix. On partisan
fairness, DualBalance reduces $|\mathrm{EG}|$ relative to the
enacted 119th-Congress plan on 14 of 20 states with precinct-level
election data (median $|\mathrm{EG}|$ 0.085 vs.\ enacted 0.133;
sign test $p = 0.058$ one-tailed). A rotation sensitivity sweep
across 12 anchor angles per state finds DBS is stable across anchors
(cross-state median $\sigma = 0.010$) while projected seat counts
shift by 1--3 seats in 26 of 41 states, identifying a pre-committed
rotation-selection rule as the primary remaining open design question.

Three limits govern adoption. Compactness scores are substantially
below enacted plans (median Polsby-Popper 0.115 vs.\ 0.281), a
structural cost of spanning the full urban-rural gradient in each
district. Where Voting Rights Act \S2 \textit{Gingles} preconditions
are met, a state-specific minority-opportunity audit is required
before implementation; the algorithm does not guarantee
minority-seat preservation. And the procedural neutrality is
symmetric: the rule that forecloses partisan engineering equally
forecloses race-conscious remediation, a trade-off for legislatures
to resolve. Within these limits, DualBalance demonstrates that a transparent,
reproducible generation rule can satisfy population-balance law
across most U.S.\ states and substantially reduce partisan asymmetry,
without reading any political data. Whether the resulting districts
would be widely accepted as fair is a separate question this paper
does not resolve. What constitutes fair representation in a
geographically diverse democracy is genuinely contested: equal
people and equal land is one principled answer, but others would
prioritize community integrity, partisan proportionality, or
geographic compactness, and reasonable people disagree on which
matters most. Some will find that spanning urban cores to rural
boundaries serves representation well; others will not. This paper
demonstrates feasibility and documents properties. Whether the
approach should be adopted is a normative and political question
the algorithm cannot answer.
\end{abstract}

\input{sections/introduction}
\input{sections/methods}
\input{sections/results}
\input{sections/discussion}

\clearpage
\appendix
\section*{Supplementary Tables}
\addcontentsline{toc}{section}{Supplementary Tables}

\input{sections/rotation_table}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
