\section{Discussion}
\label{sec:discussion}

\subsection{What the Minnesota result does and does not show}

The headline (DualBalance beats the enacted Minnesota plan on the DualBalance
Score) is a single-state existence result. It shows that radial slicing
can score competitively on the dual-balance objective without reading
party, race, or any discretionary input. It does not show that DualBalance
generalizes to all 50 states, that DBS is the right scalar to optimize,
or that DualBalance maps would survive judicial scrutiny. We take those up
below.

\subsection{Co-design of score and algorithm}

A natural objection: ``$A$ beats $B$ on Metric $M$ because $A$ and
$M$ were co-designed.'' DBS~\eqref{eq:dbs} was specified before the
algorithm choice. We tested three forms: the weighted form used here,
a sum-of-means form
$1/(1 + \overline{\mathrm{pop\_dev}} + \overline{\mathrm{area\_dev}})$,
and an $L^{\infty}$ form
$1/(1 + \max_i [0.5\,\mathrm{pop\_dev}_i + 0.5\,\mathrm{area\_dev}_i])$.
DualBalance beats the enacted plan on the first two; the enacted plan wins
on the third, because DualBalance concentrates area imbalance in a single
rural district. We use the weighted form as the headline because it
cleanly implements ``each district carries $1/N$ of people
\emph{and} $1/N$ of geography.'' Per-district deviations are reported
alongside so readers can recompute against any preferred aggregation.

The functional form satisfies four properties we wanted from the
outset. (i) \emph{Scale invariance}: the per-district inputs are
relative deviations $|x - x^*|/x^*$, so the score is invariant to
the units of population and area. (ii) \emph{Equal treatment of the
two objectives}: the $0.5/0.5$ weighting is the unique convex
combination treating $\overline{\mathrm{pop\_dev}}$ and
$\overline{\mathrm{area\_dev}}$ symmetrically. (iii)
\emph{Monotonicity}: the score strictly decreases as either
deviation increases. (iv) \emph{Boundedness in $[0,1]$}: the
reciprocal form maps zero error to $1$ and unbounded error toward
$0$, giving a calibrated reading rather than a raw cost. These do
not pin DBS down uniquely, but they constrain the design space
enough that the remaining choice (arithmetic mean of deviations
versus a max-norm or Lp aggregation) becomes a tradeoff between
average-case fairness and worst-case fairness. We report all the
per-district numbers needed to compute both.

Population density variance bounds the achievable
$\overline{\mathrm{area\_dev}}$ on any pop-balanced plan. Both DualBalance
and the enacted plan land near $\overline{\mathrm{area\_dev}} \approx
1.0$--$1.1$. On Minnesota's geometry (Twin Cities = 55\,\% of
population in 3\,\% of land area) this is a hard floor. Improvements
would require districts that span urban and rural, which is exactly
what radial slicing does. The 8.7-point gap (103.9 vs.~112.6) is the
operational measure of how much area balance DualBalance recovers given
that floor.

\subsection{Compactness is the trade}
\label{sec:discussion-compactness}

DualBalance trades compactness for area balance by design. This is not an
incidental cost we apologize for: it is the mechanism. A district
that holds $1/N$ of a high-variance density distribution must
either span the full density range (and look elongated) or sit
inside a single density regime (and produce an unbalanced area
share). Hand-drawn maps typically choose the second; DualBalance chooses
the first. The PP=0.047 worst-slice number is what choosing the
first looks like on Minnesota's geometry. The enacted Minnesota
plan's worst PP is 0.178, roughly four times higher, and its
$\overline{\mathrm{area\_dev}}$ is correspondingly 8.7 points worse.

Three observations qualify the trade.

\paragraph{Compactness is a metric, not a constitutional requirement.}
\textit{Polsby-Popper} as a measure was published in
1991~\cite{polsbypopper1991}; it appears in no federal statute and
binds no court. Courts have used compactness as \emph{evidence} of
intent in racial-gerrymandering cases (\textit{Shaw v.\ Reno} and
progeny), but the constitutional violation in those cases turns on
the use of race in line-drawing, not on the resulting shape per~se.
A deterministic, race-blind rule does not carry the racial intent that
triggers \textit{Shaw}; the resulting low compactness is a property of
the rule, not a signal of an attempt to disadvantage any group.

\paragraph{Compactness assumes a baseline of comparison.} A district
is ``bizarrely shaped'' relative to surrounding districts that are
not. If all eight districts in a state are produced by the same
geometric rule and all eight share a similar slice-like character,
the comparative baseline shifts. The aesthetic objection to long thin
shapes is partly historical: it picks out districts that depart from
the surrounding norm of compact blobs. A scheme in which every
district is a slice does not pick anything out.

\paragraph{The Voting Rights Act trade-off is real.}
\textit{Allen~v.\ Milligan}~\cite{milligan2023} reaffirmed Section~2's
effects-based vote-dilution doctrine; \textit{Alexander}~\cite{alexander2024}
and \textit{Callais}~\cite{callais2026} narrowed its practical reach.
Section~2 asks about effects on minority voters' opportunity to elect,
not about intent. DualBalance is race-blind, but blindness to race does not
satisfy Section~2 by itself. In jurisdictions where the
\textit{Gingles}~\cite{gingles1986} preconditions are met, DualBalance may
produce a plan that fails Section~2 review even though no one drew
lines to dilute. The legislative question (whether to accept reduced
minority opportunity in exchange for a generator that cannot be
partisan-tuned) is for legislatures and courts, not for the
algorithm.

\subsection{Relationship to prior deterministic methods}
\label{sec:discussion-prior}

DualBalance is, to our knowledge, the only deterministic districting
method whose objective is bivariate (population \emph{and} area). Every
other published deterministic method optimizes a single extensive
criterion -- usually population balance plus a geometric proxy
(compactness or shortest-cut length) -- and that single-axis design
is empirically why none of them clears all four of the criteria we
set out in the introduction: legal viability under \emph{Karcher},
DBS competitiveness, partisan-fairness on the gerrymandered states,
and minority-majority district creation where the geography supports
it. We argue this point empirically for the two methods we benchmark
on all 50 states (Cascade and BDistricting) and structurally for the
remainder.

\paragraph{Cascade (Iowa-LSA-flavored, county aggregation).}
Cascade prioritizes county integrity lexicographically:
aggregate VTDs to counties, capacitated first-fit
counties to districts, split a county only when its population
exceeds the per-district cap. The county-integrity priority makes
the algorithm legitimate \emph{in Iowa}, where no county is large
enough to dominate a district, but on every state with a major
metropolitan county it produces population deviations far above
\emph{Karcher}: 0.50\,\% on Wisconsin, 10.27\,\% on North Carolina,
24.58\,\% on Texas, 41.56\,\% on Massachusetts, 76.14\,\% on
Minnesota (Table~\ref{tab:multistate-dbs}). Cascade is therefore
\emph{not legally viable} on the majority of states, regardless of
how well it does on DBS or efficiency gap. Treating Cascade as a
counterfactual `national rule' would constitutionally disqualify
the resulting maps under \emph{Reynolds v.~Sims}, let alone
\emph{Karcher}.

\paragraph{BDistricting~\cite{olsonbd}, representing the
capacitated-$k$-means / centroidal-power-diagram
family~\cite{cohenaddad2018,levinfriedler2019}.} BDistricting
minimizes population-weighted Euclidean distance from each unit to
its district center subject to equal-population caps, with
Lloyd-style iteration to convergence -- an empirical realization of
the centroidal-power-diagram objective formalized by
Cohen-Addad, Klein, and Young~\cite{cohenaddad2018} and
independently studied by Levin and Friedler~\cite{levinfriedler2019}.
The objective is unidimensional: equal population plus geometric
compactness around dispersed centers. In our six-state comparison
BDistricting achieves the highest Polsby-Popper and Reock scores of
any algorithm on every state but is uneven on population (\emph{Karcher}
on some states, marginal on others) and middle-of-the-road on DBS
because area balance is not part of the objective. The general
pattern: a deterministic generator that optimizes
population~$+$~compactness with no third axis cannot reliably hit
the dual-balance objective, because the radial mixing of dense and
sparse territory that gives DualBalance its area balance is
explicitly destroyed by Lloyd recentering, which pulls seeds toward
their cells' centers of mass.

\paragraph{Shortest Splitline~\cite{smithsplit}.} Smith's algorithm
recursively bisects the state with the shortest
population-balancing line. The objective is unidimensional in a
different sense: at each recursive step it minimizes one number
(line length) and otherwise lets the geometry fall where it
will. The resulting districts are often visibly elongated, with
boundaries that follow neither county lines nor visible geographic
features. No U.S.~jurisdiction has adopted Splitline; no public
50-state baseline exists; the algorithm is exemplary as a proof of
deterministic feasibility but not as a deployment candidate. We
include it here for completeness rather than for empirical
comparison.

\paragraph{Cascade as our chosen empirical baseline.} We include
Cascade in the comparison rather than Splitline because Cascade is
the deterministic alternative that has actually been deployed at
scale (in Iowa since 1980) and is the most-cited model of impartial
districting in the popular literature. The empirical contribution
of the comparison is to show that the Iowa template is geographically
contingent: on the 49 other states, Cascade fails the legal-viability
test catastrophically. The conceptual contribution is to clarify
that being deterministic and impartial is necessary but not
sufficient. The objective the algorithm pursues matters, and the
DualBalance objective -- equal population \emph{and} equal area -- is
both expressible deterministically and broad enough to admit
legally-viable plans on every state we have tested.

The forensic critique in \S\ref{sec:discussion-forensic} applies
equally to all of these methods, and to any future deterministic
generator built on similar principles: the gerrymandering metrics
were built to investigate human line-drawers, and on any
deterministic plan their inferential power evaporates.

\subsection{Limitations}

\paragraph{VTD data availability.} The Census Bureau's TIGER 2020PL
redistricting file includes VTD boundaries only for states that
submitted them. At least three states --- California, Hawaii, and Oregon ---
did not participate in the VTD program for 2020 and therefore have no
\texttt{vtd20} shapefile in TIGER; they are excluded from all
cross-state tables and figures here. Future work can substitute
block-group or block boundaries, which the Census does publish for all
states, at a cost of larger input size and longer optimizer runtime.

\paragraph{Peninsula and polycentric geometries.} DualBalance's optimizer
can fail to close the Karcher gap in states whose geometry defeats
single-center radial seeding. Florida is the clearest case in our
validation set: with 28 seats distributed across a long peninsula,
the radial slices created around the population-weighted centroid
produce districts that the VTD-scale optimizer cannot balance to
the 0.05\,\% target in $10^5$ passes, leaving
$\mathrm{pop\_dev}_{\max} = 44.2\,\%$. The block-scale refinement
pass (\S\ref{sec:methods-block}) is the next step for these
hard-geometry states; results will be reported as they are
available.

\paragraph{Expanding evidence base.} Tables and figures in
\S\ref{sec:results-multistate} will be updated as the full-state
harness completes. Figures and tables note $\dagger$ for any state
where VTD data were unavailable and results therefore could not be
computed.

\paragraph{Worst-district area outlier.} DualBalance concentrates the
area imbalance in a single rural district (here, District~2 at
275\,\% over-target). This is geometric, not algorithmic: any
pop-balanced partition of a state with Minnesota's density profile
must give some rural district disproportionate area. DualBalance's
$\mathrm{area\_dev}_{\max}$ of 275\,\% is worse than the enacted
plan's 241\,\% even though the mean is lower, so DualBalance distributes
the imbalance unevenly. Whether evenly distributed area imbalance is
preferable to a single concentrated outlier is a separable normative
question.

\paragraph{No multi-unit transportation step.} The optional
\texttt{--tighten-pop} pass is a greedy single-unit swap. A true 2D
transportation LP bounding both population and area would directly
minimize the DBS objective; we leave that to future work.

\paragraph{Computational cost.} Pure DualBalance runs in under a second on
Minnesota. The tightening pass takes $\sim$18~s for 80 swaps. We have
not profiled California or Texas; worst-case scaling is $O(|U|^2 N)$
per swap due to the contiguity check.

\paragraph{The compactness floor.} The worst slice's
$\mathrm{PP}=0.047$ is well below the informal court-testimony
threshold of 0.10. We argue in \S\ref{sec:discussion-compactness}
that the \emph{Shaw}~\cite{shaw1993} doctrine turns on intent rather
than shape per se, but readers who reject that argument should
regard the 0.047 figure as a hard ceiling on legal viability.

\subsection{Future work}
\label{sec:future-work}

\paragraph{Broad-coverage validation (ongoing).} The six-state
initial validation (MN, IA, MA, TX, NC, WI) is being extended to
the full set of multi-seat states for which TIGER 2020PL VTD
boundaries are available (41 states at time of writing; CA, HI,
OR excluded). Results will be incorporated into the published
version of this paper. The most analytically important remaining
states are Pennsylvania and Ohio (state-court partisan-gerrymandering
jurisprudence) and the four states where DualBalance is still above
the Karcher threshold at VTD scale (FL in particular, where the
peninsula geometry makes single-center radial seeding hard to
balance).

\paragraph{Multi-center radial seeding.} For polycentric states, place
seeds on circles around \emph{each} of $k$ population centers, with
$k$ chosen deterministically from the density distribution. Not yet
implemented.

\paragraph{Direct 2D transportation step.} Replace the
DualBalance-plus-tighten two-stage pipeline with a single optimization that
minimizes DBS under contiguity constraints. More expensive than the
current pipeline; the mathematically clean answer to ``which plan
directly minimizes the DualBalance objective.''

\paragraph{Splitline benchmark.} The six-state benchmark
(\S\ref{sec:results-multistate}) covers DualBalance, Cascade, BDistricting,
and the enacted plan. Adding Smith's Shortest
Splitline~\cite{smithsplit} would give a fourth deterministic
baseline; computing it from scratch on each state is straightforward
but has not yet been done.

\paragraph{Scoring-harness extensions.} The current harness reports
population, area, Polsby-Popper, and Reock. Natural additions include
partisan diagnostics (efficiency gap, mean-median, declination) on
voter-registration data, county-split counts, and minority
representation diagnostics under Section~2. These are reporting
extensions and would not feed back into the generator
(\S\ref{sec:methods}, ``out-of-scope inputs'').

\subsection{Forensic metrics applied to a generative procedure}
\label{sec:discussion-forensic}

Two problems arise when forensic metrics are pressed into service as
design objectives rather than as diagnostic tools. First, optimizing
for a plan that ``looks like what a neutral process would have
produced'' is a moving target: the counterfactual shifts with the
ensemble distribution one chooses and the legal constraints one
encodes, both of which carry discretionary content~\cite{cain2014}.
Second, fully content-neutral automated procedures can produce
systematic partisan asymmetry as a geographic artifact. Chen and
Rodden~\cite{chenrodden2013} showed that residential clustering
causes impartial automated maps to disadvantage Democrats in many
states, so a nonzero Efficiency Gap on such a map reflects
geography rather than intent. Forensic metrics were designed to
detect manipulation by a human line-drawer; absent a line-drawer,
their inferential content evaporates~\cite{pildes2004}.

Deterministic generation has interpretive consequences for the
fairness-metrics literature. Most gerrymandering diagnostics carry
two functions. They \emph{describe} a property of the plan (a shape,
a vote distribution, a count of split counties). And they support an
\emph{inference} about the line-drawer's motive (that the shape was
chosen, that the votes were packed and cracked, that the splits
followed a partisan or racial pattern). The two are nearly always
entangled in practice because the literature was built around
enacted plans, where intent is always at issue.

DualBalance separates them. The descriptive function survives intact: a
$\mathrm{PP}$ of 0.09 still names a non-compact shape, with
administrative and legitimacy consequences; an Efficiency Gap of
$+6.6\,\%$ still names a real wasted-vote asymmetry, with
representational consequences; zero majority-minority districts
still names a partition under which no minority group reaches a
50\,\% VAP share, with Section~2 consequences. The inferential
function does not. DualBalance is a pure function of geometry, population,
and seat count, so the chain from observed property back to motive
has no actor to terminate on.

Three implications, narrowed accordingly.

\paragraph{\emph{Shaw}-style shape inference loses purchase, but
compactness still matters administratively.}
\textit{Shaw v.~Reno}~\cite{shaw1993} uses bizarre shape as evidence
that race predominated in line-drawing. The inference chain runs
shape $\to$ unusual process $\to$ illegitimate criterion $\to$ race.
On DualBalance the first link is broken: every district produced by the
rule has the same shape character, so unusual shape no longer
implies unusual process. \emph{Shaw} does not reach a race-blind
algorithm. But compactness still has a non-inferential role.
Districts that are hard to identify with on a map, hard to administer,
or hard to campaign in carry costs independent of who drew them.
DualBalance's PP=0.047 worst slice is a real cost. The argument here is
not that compactness ceases to matter, only that it ceases to be
evidence of motive.

\paragraph{Partisan metrics describe effects, not intent.}
\textit{Rucho v.~Common Cause}~\cite{rucho2019} foreclosed federal
partisan-gerrymandering review for lack of a manageable standard for
distinguishing acceptable from unacceptable partisan motivation. The
question is moot on DualBalance (there is no motivation), but the partisan
\emph{effect} the metrics measure is not. An Efficiency Gap of
$+6.6\,\%$ on a DualBalance map still corresponds to wasted-vote asymmetry
that affects representation; state-court tests that interpret EG as
evidence of effect, rather than as evidence of intent, still bind.
The rhetorical move from ``the map shows EG $+6.6\,\%$'' to
``someone built that EG $+6.6\,\%$'' becomes unavailable on DualBalance,
but the metric continues to measure what it measured.

\paragraph{Ensemble outlier tests lose their baseline.} The Duke and
MGGG ensembles~\cite{mattinglyduke,recom2021} compare an enacted plan
to a sample of legally compliant alternatives a neutral line-drawer
might plausibly have produced. The sample stands in for the absent
neutral line-drawer; the enacted plan being an outlier suggests the
actual line-drawer differed from the neutral one. DualBalance \emph{is}
the neutral line-drawer in a stronger sense than the ensemble's
sample mean. Asking whether the DualBalance plan is an outlier against its
own ensemble may not be meaningful: the ensemble collapses to one
point. Ensembles can still play a comparison role across
\emph{different} algorithmic generators (e.g.\ DualBalance vs.\ BDistricting
vs.\ splitline), but the question is no longer ``is this map
suspiciously different from neutral?''

Metrics remain informative as descriptive summaries of plan effects.
A DualBalance plan with $\mathrm{PP}_{\min} = 0.047$, Efficiency Gap
$+6.6\,\%$, and zero majority-minority districts has low compactness,
an R-favoring wasted-vote asymmetry, and no MMD. Those facts carry
administrative, representational, and Section~2 consequences. The
inference from those facts to a person who chose them is what
deterministic generation removes.

\subsection{What this paper claims, and what it does not}

The paper makes four claims.

First, that a knob-free, race-blind, partisan-blind, deterministic
districting algorithm exists that produces output competitive with,
and on a single empirically tested state superior to, the
corresponding hand-drawn enacted plan on a prespecified
population-and-area objective. The algorithm's output is
byte-reproducible and updates only with the ten-year census.

Second, that the algorithm's procedural neutrality is symmetric. The
rule cannot be used to harm any group, party, or incumbent, and
cannot be used to help any of them. The same property that forecloses
a partisan gerrymander forecloses a race-conscious remedy. The
symmetry is the algorithm's defining property; whether it is morally
preferable to the status quo of partisan and remedial discretion is
a question for legislatures, courts, and voters.

Third, that the post-\textit{Callais} legal landscape gives a
generator of this kind a narrower constitutional exposure than any
race-conscious or partisan-conscious alternative. Federal partisan
review is foreclosed under \textit{Rucho}; federal racial-gerrymander
risk is contracting under \textit{Alexander} and \textit{Callais};
state-court remedies for partisan claims depend on the politics of
each state's supreme court. A rule whose inputs do not contain race
or party is, in the strict sense, not subject to either body of
doctrine.

Fourth, the conventional gerrymandering metrics (compactness scores,
Efficiency Gap, mean-median, ensemble outliers, majority-minority
counts) play two roles on enacted plans, only one of which transfers
to DualBalance. As descriptions of plan effects (shape, wasted-vote
asymmetry, minority opportunity) they remain valid and are reported
alongside the DualBalance Score. As evidence of line-drawer intent
they do not transfer. The intent-attribution use of these metrics,
not their measurement use, is what the generative case breaks. This
is the most contested framing claim in the paper.

The paper does not claim that DBS is universally the right objective,
that DualBalance dominates enacted plans on all states, that compactness
can be ignored, that effects-based fairness analyses (including
Section~2) become irrelevant, or that the metric toolkit developed
over the past thirty years is wrong. The narrow claim is that the
intent reading of those metrics presupposes a line-drawer who is
not present in this construction.

Whether deterministic generation is preferable to discretionary
redistricting is a political and constitutional question; the
algorithm itself cannot answer it.
