\section{Introduction}

This paper is principally about how districting metrics should be
read once line-drawing becomes a deterministic function rather than
a discretionary act. The algorithm we propose, PRISM, is the
construction that makes the question concrete; the legal-doctrinal
and metric-interpretation arguments that follow are where the
paper's main weight sits.

\subsection{The Instability of Human-Drawn Districts}

Legislative redistricting in the United States is structurally unstable. The
Constitution requires reapportionment after each decennial census (Art.~I, \S2)
but specifies neither a method for allocating seats among the states nor a
procedure for drawing district boundaries within them. Both decisions have
been delegated, by default, to political actors. The result, predictable from
the design, is that maps are routinely drawn to favor whichever group controls
the pen at the moment of redistricting. The Supreme Court has acknowledged the
problem in its strongest terms while declining to remedy it: in
\emph{Rucho v.~Common Cause}~\cite{rucho2019}, Chief Justice Roberts wrote for
a 5--4 majority that partisan gerrymandering claims present nonjusticiable
political questions, conceding that the practice is ``incompatible with
democratic principles'' but holding that no manageable judicial standard
exists. Justice Kagan's dissent, joined by Ginsburg, Breyer, and
Sotomayor, argued the opposite: that lower courts had already converged on
workable tests, and that ``of all times to abandon the Court's duty to declare
the law, this was not the one''~\cite{rucho2019}.

Race-conscious districting operates under a separate and increasingly
constrained regime. Section~2 of the Voting Rights Act, as construed in
\emph{Thornburg v.~Gingles}~\cite{gingles1986}, requires majority-minority
districts where the three Gingles preconditions are met; the Fourteenth
Amendment, as construed in \emph{Shaw v.~Reno}~\cite{shaw1993} and its
progeny, subjects to strict scrutiny any map in which race ``predominates''
in the line-drawing. The two doctrines pull in opposite directions. The
Supreme Court reaffirmed Section~2 in \emph{Allen v.~Milligan}~\cite{milligan2023},
requiring Alabama to draw a second majority-Black district. One year later,
in \emph{Alexander v.~South Carolina Conf.~of the NAACP}~\cite{alexander2024},
the Court reversed a district court's racial-gerrymander finding and raised
the plaintiff's evidentiary burden, holding that challengers must
``disentangle race from politics'' and produce illustrative alternative maps
that satisfy all of the state's legitimate goals. In April 2026, in
\emph{Louisiana v.~Callais}~\cite{callais2026}, the Court further held that
even a map drawn to comply with a prior \emph{Section~2} order may itself
violate the Equal Protection Clause if the VRA did not in fact compel the
race-based remedy. The combined effect of \emph{Alexander} and
\emph{Callais} is that race-conscious line-drawing now faces a narrower
window of constitutional safety than at any point since the VRA's adoption.

Two structural consequences follow. First, the only federal channel for
gerrymandering review is racial; partisan claims are foreclosed under
\emph{Rucho}. Second, because Black voters in the United States vote
roughly nine-to-one Democratic, race and partisanship are statistically
entangled~\cite{hasen2018,coluraceparty}. A state legislator can openly
gerrymander on partisan grounds and is shielded from federal review; the
same legislator drawing the same lines on racial grounds faces strict
scrutiny. After \emph{Rucho}, state constitutional review under
\emph{Moore v.~Harper}~\cite{moore2023} remains the only judicial avenue
for partisan claims, but it depends on the politics of each state's
supreme court and the language of its state constitution. Roughly ten
state supreme courts have so far recognized state-constitutional
partisan-gerrymandering claims; the rest have not~\cite{stateline2025}.

Federal partisan review is foreclosed. Section~2 has been narrowed.
State review is uneven and politically contingent. Redistricting has
shifted from a once-per-decade event into a continuous partisan
exercise: several states have already redrawn maps mid-decade in
response to electoral results rather than census revision, and the
practice is expected to expand. A regime that redraws lines
whenever the pen changes hands provides neither stable
representation nor predictable constituencies.

\subsection{The Normative Paradox of ``Bias Detection''}

In response to \emph{Rucho}'s manageability concern, a substantial
mathematical literature has produced quantitative tests for partisan
gerrymandering. The Efficiency Gap of Stephanopoulos and
McGhee~\cite{stephanopoulosmcghee2015} measures the disparity in
``wasted votes'' between parties and was central to the lower-court
ruling in \emph{Whitford v.~Gill}. Wang's three tests~\cite{wang2016},
namely mean--median difference, lopsided wins, and a $t$-test for
partisan asymmetry, played a similar role in \emph{LWV v.~Commonwealth} of
Pennsylvania. Warrington's declination~\cite{warrington2018} measures the
asymmetry of the seats--votes kink at 50\%. Most influentially, the
Markov Chain Monte Carlo ensemble methods developed at Duke
University~\cite{mattinglyduke} and the MGGG Redistricting
Lab~\cite{recom2021} construct large samples of legally compliant
neutral maps and ask whether an enacted plan is an outlier with respect
to that distribution. Ensemble methods have been admitted as evidence
in \emph{Common Cause v.~Lewis}~(N.C.~2019), \emph{LWV v.~Commonwealth}
(Pa.~2018), and as plaintiff support in \emph{Allen v.~Milligan}.

Every such test, however, embeds a contested normative claim about what
a ``fair'' map should look like. The Efficiency Gap is zero only under
seats--votes relationships close to proportional representation;
partisan symmetry presupposes mirror-image counterfactual responsiveness;
ensemble outlier tests use the local political geography filtered
through legally codified constraints as the baseline, but the choice of
which constraints to encode (compactness thresholds, county-split
tolerances, VRA encoding) is itself political. Chen and
Rodden~\cite{chenrodden2013} demonstrated that in many states even
content-neutral automated maps produce systematic seat bias against
Democrats because of residential clustering, so a nonzero Efficiency Gap
or mean--median value may reflect geography rather than intent. Pildes
has long argued that no neutral baseline exists~\cite{pildes2004}, and
Cain has framed districting as a tradeoff among incommensurable goods
that no scalar metric can collapse without bias~\cite{cain2014}.

Beneath the surveyed objections sits a sharper one about what these
metrics are \emph{for}. Each plays two roles in the literature:
a descriptive role (the metric measures an actual property of the
plan: shape, wasted votes, partisan-share distribution) and an
inferential role (the metric serves as evidence of the line-drawer's
intent). The two roles are usually entangled in practice because
the literature was built around enacted plans, where intent is
always at issue.

Polsby-Popper~\cite{polsbypopper1991} and Reock~\cite{reock1961}
flag shapes too contorted to have arisen by ordinary line-drawing;
the Efficiency Gap~\cite{stephanopoulosmcghee2015}, mean-median
asymmetry~\cite{wang2016}, declination~\cite{warrington2018}, and
the Duke and MGGG ensemble outlier
tests~\cite{mattinglyduke,recom2021} flag partisan profiles that
look engineered. The inferential power of all of these depends on
there being a line-drawer to investigate.

Remove the line-drawer and the descriptive role survives intact. A
radial slice with $\mathrm{PP} = 0.09$ is still geometrically
non-compact, with all the administrative and legitimacy implications
that follow. An Efficiency Gap of $+6.6\,\%$ on a deterministic plan
still describes a real asymmetry in wasted votes, with real
representational consequences. What the metrics no longer do, on a
plan with no line-drawer, is support the inference back to motive.
Section~2 of the Voting Rights Act, which asks about
\emph{effects}, still applies. \emph{Shaw}-style intent inference,
which asks who drew the line and why, does not.

\subsection{Deterministic Districting: A Sixty-Year Tradition}

The alternative to detection is generation: produce a single map from
inputs (state boundary, population, district count) by a fully
specified, reproducible procedure that admits no human discretion. The
intellectual line begins with Vickrey~\cite{vickrey1961}, who proposed
in 1961 that boundary line-drawing be done by ``an automated and
impersonal procedure'' to remove the discretion gerrymandering
exploits. Hess et~al.~\cite{hess1965} formalized the idea four years
later as a capacitated transportation problem, alternating between an
assignment step (minimize the population-weighted sum of squared
distances from each unit to a district center subject to equal
population) and an update step (recompute each center as the
population-weighted centroid of its assigned units). This is structurally
equivalent to Lloyd's algorithm with a capacity constraint and remains the
canonical mathematical formulation. Mehrotra, Johnson, and
Nemhauser~\cite{mjn1998} subsequently cast the problem as a column-generation
set partitioning program admitting exact optimization at moderate
scales.

Two contemporary lines extend this work. Brian Olson's
BDistricting~\cite{olsonbd} is a publicly available implementation
that minimizes population-weighted distance to district centers with
equal-population constraints, applied at census-block resolution; maps
have been generated for every state and updated each decennial census.
Cohen-Addad, Klein, and Young~\cite{cohenaddad2018} provide the
rigorous reformulation as balanced \emph{centroidal power diagrams},
generalizing the Voronoi tessellation by assigning each seed an
additive weight (the Aurenhammer~\cite{aurenhammer1987} construction)
and solving for weights that enforce equal mass per cell.
Levin and Friedler~\cite{levinfriedler2019} independently rediscovered
the capacitated $k$-means formulation and applied it to all fifty states.
A parallel branch, Warren Smith's Shortest Splitline
algorithm~\cite{smithsplit}, recursively bisects the state with the
shortest population-balancing line; it is fully deterministic and
unique but lacks any global compactness or boundary-awareness criterion.

Three observations link this tradition to the current work. First, all
extant deterministic algorithms balance \emph{one} extensive quantity
across districts, namely population, and treat geographic area, if at all,
only indirectly through compactness shape metrics such as Polsby-Popper~\cite{polsbypopper1991}
or Reock~\cite{reock1961}. Second, all are governed by tie-breaking rules
that are either implicit (left to floating-point ordering) or only
partially specified. Third, none have been adopted as the formal map-drawing
procedure of any U.S.~jurisdiction, though Iowa's Legislative Services
Agency~\cite{iowalsa} operates under a strict criteria cascade that has
produced widely accepted maps since 1980.

\subsection{Contribution: DualBalance Districting and the PRISM algorithm}

The U.S. Constitution divides representation along two axes: the
House apportions seats by population (Art.~I, \S2), the Senate by
sovereign state regardless of population (Art.~I, \S3). Madison's
defense in \emph{Federalist} Nos.~54--58~\cite{federalist54}
treats this as a principled refusal to make representation reduce
to a single dimension. We do not claim that the Senate represents
land area; it represents states. We take the framers' refusal as
motivating intuition: within a single chamber, district lines
should not collapse representation to population alone.

\emph{DualBalance Districting} operationalizes that intuition with
two coequal targets. Each district should carry roughly $1/N$ of a
state's people \emph{and} roughly $1/N$ of its land:
\begin{equation}
  \mathrm{DBS}
    = \frac{1}{1
       + \tfrac{1}{2}\,\overline{\mathrm{pop\_dev}}
       + \tfrac{1}{2}\,\overline{\mathrm{area\_dev}}},
\end{equation}
with $\mathrm{pop\_dev}_i = |P(D_i) - P^{*}|/P^{*}$ and
$\mathrm{area\_dev}_i = |A(D_i) - A^{*}|/A^{*}$,
$P^{*} = P/N$, $A^{*} = A/N$, overlines denoting means. The area
term is not Senate-style sovereignty; it is the
operationalization of a simpler claim: districts that span both
metropolitan and rural territory force each elected representative
to answer to constituents across the state's density range, rather
than to a single dense metro or a single rural hinterland. Its
discussion of trade-offs against compactness and minority
representation appears in \S\ref{sec:discussion-compactness} and
\S\ref{sec:discussion-forensic}.

\emph{PRISM} (Population-weighted Radial Impartial Slicing Method)
is the algorithm that pursues this objective. PRISM places $N$ seeds
in a small circle around the state's population-weighted centroid and
assigns each census unit to the nearest seed with capacity remaining.
The districts come out as radial slices through the population center.
Each slice spans both dense (near-center) and sparse
(boundary-side) territory, so the area each slice inherits trends
toward $A^{*}$ by geometry rather than by penalty. The pipeline (seed
placement, capacitated assignment, contiguity repair) is a pure
function of $(\text{units}, N)$: no random seed, no iteration, no
tunable weight. Identical inputs always yield byte-identical outputs.
A single optional post-pass (\verb|--tighten-pop|) closes the residual
per-district population gap to a user-supplied \emph{Reynolds
v.~Sims} tolerance via greedy $L^{1}$ boundary swaps. Apportionment
across states uses the Method of Equal Proportions, the standard since
1941~\cite{balinskiyoung}. Updates occur only with the decennial
census.

The procedural neutrality is symmetric. PRISM cannot harm any group,
party, or incumbent; it equally cannot help any of them. The same
property that forecloses a partisan gerrymander forecloses a
race-conscious remedy.

The legal exposure is correspondingly narrow. After
\emph{Rucho}~\cite{rucho2019}, federal partisan review is foreclosed.
After \emph{Alexander}~\cite{alexander2024} and
\emph{Callais}~\cite{callais2026}, a generator that does not consider
race cannot be a racial gerrymander under \emph{Shaw}~\cite{shaw1993}.
State-court partisan tests rely on outcome metrics (Efficiency Gap,
mean--median, ensemble outliers) that a geography-only generator
should pass routinely, with the caveat that those metrics already
measure geography on enacted plans
too~\cite{chenrodden2013}. A residual risk is Section~2: in
jurisdictions where the Gingles preconditions are met, a race-blind
plan may not satisfy Section~2 where a race-conscious plan would.

\paragraph{Contributions.}
\begin{enumerate}
  \item PRISM: a deterministic, knob-free districting algorithm
        producing radially-sliced districts that balance population
        and area equally. Pure function of $(\text{units}, N)$.
  \item A fully specified tie-breaking cascade guaranteeing
        byte-identical output, plus a proof that the cascade resolves
        every comparison the algorithm makes.
  \item A scoring harness decoupled from the generator, applying the
        same metrics to PRISM, enacted plans, and alternatives.
        Empirical results on Minnesota show PRISM beating the enacted
        119\textsuperscript{th}-Congress map on the DualBalance Score.
  \item An argument that the standard gerrymandering metrics
        (compactness, Efficiency Gap, mean--median, ensemble outliers,
        majority-minority counts) are forensic instruments built to
        infer human intent, and so do not license, on a PRISM plan,
        the inferences they license on an enacted plan.
\end{enumerate}

Section~2 formalizes PRISM. Section~3 reports the Minnesota result.
Section~4 discusses limitations and legal considerations.
