\section{Methods}
\label{sec:methods}

\subsection{Data and inputs}
\label{sec:methods-data}

Atomic units are U.S.\ Census Bureau \emph{voting tabulation
districts} (VTDs) and \emph{census tabulation blocks} from the 2020
decennial release~\cite{census2020pl}, obtained via TIGER/Line
shapefiles~\cite{tiger2020}. Boundaries are used as-is; no
simplification or smoothing is applied. Population counts are the
total-population figures from PL~94-171. The enacted congressional
plan used for comparison is the 119th-Congress plan (effective
January 2025), joined to VTDs and blocks via a spatial
representative-point join against the TIGER 2024 \texttt{cd119}
layer~\cite{tiger2024cd119}.

\paragraph{Note on enacted-plan population deviations.}
Enacted U.S. congressional plans are litigated to \textit{Karcher}
compliance at the census-block level. The enacted-plan deviations
reported in this paper reflect the VTD-layer spatial join: a VTD
that straddles a district boundary is assigned in full to one
district, introducing apparent deviation where the legislative
record shows none. DualBalance's reported deviations come from its
own block-refined output and are directly comparable to the
\textit{Karcher} threshold. Counts of enacted-plan Karcher compliance
(Tables~\ref{tab:multistate-dbs}--\ref{tab:aggregate-comparison})
should be read as a conservative lower bound on enacted compliance,
not as legally cognizable deviations: the true enacted deviations
are near zero on block-accurate data.

\paragraph{Reproducibility.}
The \texttt{dualbalance} Python package and per-state configuration
files reproduce all tables and figures from publicly available
TIGER and Census source files.
Algorithm outputs are byte-identical across repeated runs on the
same input data; the \texttt{votes\_source} field in each state's
units file records which election data source was used.

\subsection{Problem statement}

Given a state $S$ partitioned into atomic units
$U = \{u_1, \ldots, u_M\}$ (census tabulation blocks, block groups,
or voting tabulation districts (VTDs), in increasing order of size) and
a target district count $N$, DualBalance Districting produces a
deterministic assignment $\pi : U \to \{1, \ldots, N\}$ such that
each district $D_i = \pi^{-1}(i)$ is contiguous, non-empty, and as
close as the geometry allows to representing both $1/N$ of the
state's people and $1/N$ of its geography. Let
$P = \sum_u \mathrm{pop}(u)$ and $A = \sum_u \mathrm{area}(u)$ with
per-district targets $P^* = P/N$ and $A^* = A/N$.

The full pipeline runs in three stages: a radial-seed stage,
\emph{DualBalance} (\S\S\ref{sec:methods-seeds}--\ref{sec:methods-repair}),
which provides the deterministic starting configuration; a
VTD-scale tightening stage (\S\ref{sec:methods-optimize}) which
drives $\max_i |\delta_i|/P^{*}$ toward the user-supplied tolerance
$\tau$ via a hybrid greedy local search with bounded-chain escape;
and a block-scale refinement stage (\S\ref{sec:methods-block})
which re-initializes at finer granularity to recover area balance
within the achieved pop-deviation envelope. Each stage is a pure
function of its inputs.

\subsection{Radial seed placement}
\label{sec:methods-seeds}

Compute the population-weighted centroid
\[
  c = (c_x, c_y) =
  \left(
    \frac{\sum_u p_u\,x_u}{\sum_u p_u},\;
    \frac{\sum_u p_u\,y_u}{\sum_u p_u}
  \right),
\]
where $x_u, y_u$ are the centroid coordinates of unit $u$ in an
equal-area projection and $p_u$ its population. Let $\mathrm{diag}$
denote the bounding-box diagonal of the units and set the seed
radius $r = 0.001 \cdot \mathrm{diag}$. For $d = 0, 1, \ldots, N-1$
place seed $s_d$ at
\[
  s_d = \bigl(c_x + r\cos(2\pi d/N),\, c_y + r\sin(2\pi d/N)\bigr).
\]
Seed 0 sits due east of the centroid; seeds advance counter-clockwise
by equal angular steps. The choice $r = 0.001 \cdot \mathrm{diag}$ is
small enough that the Voronoi cells generated by the seeds degenerate
to near-perfect radial slices through $c$, but large enough to keep
the seed positions numerically distinct.

These angular parameters, seed 0 due east and counter-clockwise
advance, are pre-committed design choices fixed before any
state-specific data are examined. Different anchor angles
$\theta \in [0,\,2\pi/N)$ produce different partitions with
different partisan and demographic properties, because the resulting
radial slices land differently on the state's population geography.
Rotation sensitivity, i.e., the distribution of DBS, $|\mathrm{EG}|$,
and seat outcomes across the full rotation range, is therefore an
important robustness check; results are reported in
\S\ref{sec:results-rotation}. The due-east anchor was selected as
the natural zero of the standard angular convention; an adopting
body would need to pre-specify a rotation rule (or adopt this one)
as part of any legislative implementation.

The radial configuration is what carries the dual-balance property:
each slice spans both the dense (near-$c$) and sparse (boundary-side)
territory of the state, so the population it inherits is bounded by
the cap (\S\ref{sec:methods-assign}) while the area it inherits is
driven toward $A^*$ by the slicing geometry.

\subsection{Capacitated first-fit assignment}
\label{sec:methods-assign}

Let $d(u, i) = \|x_u - s_i\|$ be the Euclidean distance from unit $u$
to seed $i$. Initialize remaining capacities $\rho_i = P^*$ for
$i = 1, \ldots, N$. Sort all $(u, i)$ pairs by ascending normalized
distance $d(u, i) / \mathrm{diag}$ and walk the sorted list in order:
\[
  \begin{array}{l}
    \text{for each } (u, i)\text{ in ascending } d(u,i)/\mathrm{diag}: \\
    \quad \text{if $u$ already assigned: continue} \\
    \quad \text{if $\rho_i \geq p_u$: assign $u$ to $i$;\;\; }\rho_i \leftarrow \rho_i - p_u \\
    \quad \text{else: skip}
  \end{array}
\]
Population balance is enforced as a hard cap: no district receives
more than $P^*$. Any unit not placed by the end of the walk (a rare
integer-rounding edge case) is assigned to the district with the
largest remaining capacity; \texttt{argmax} resolves ties to the
smallest district id.

Ties in normalized distance break by ascending
$(\mathrm{unit\_id},\,\mathrm{district\_id})$. There is no Lloyd
recentering, no iteration count, no convergence test: the radial
seeds do not drift, so a single assignment pass suffices.

\subsection{Contiguity repair}
\label{sec:methods-repair}

After \S\ref{sec:methods-assign} every unit belongs to exactly one
district, but a district may consist of more than one connected
component (rare on convex states; more common on those with
peninsulas, islands, or rural enclaves). For each such district,
the largest connected component by total population is retained;
units in the smaller components are reassigned to neighboring
districts one at a time, in the lowest-cost direction. Cost is
\[
  c(u, j) = \frac{d(x_u, s_j)}{\mathrm{diag}}
          + \frac{|P(D_j) + p_u - P^*|}{P^*}
          + \frac{|A(D_j) + a_u - A^*|}{A^*},
\]
combining a normalized distance term with normalized population and
area deviation penalties for the receiving district. Ties break in
cascade $(c, \mathrm{pop\_pen}, \mathrm{area\_pen}, \mathrm{dist},
\mathrm{district\_id})$ ascending. The repair sweep iterates until no
district has more than one connected component, capped at ten sweeps;
in practice it converges in zero or one sweep on real census
geometries.

Algorithm~\ref{alg:dualbalance} summarizes the full core pipeline.

\begin{algorithm}[htbp]
\caption{DualBalance Districting}
\label{alg:dualbalance}
\begin{algorithmic}[1]
\Require Atomic units $\mathcal{U}=\{u_1,\dots,u_M\}$ with populations
         $\{p_u\}$, centroids $\{x_u\}$, and areas $\{a_u\}$;
         district count $N$; population tolerance $T$
\Ensure  Assignment $\pi:\mathcal{U}\to\{1,\dots,N\}$;
         every district contiguous, non-empty, with
         $\max_i|\delta_i|/P^* \leq T$
\vspace{4pt}
\State $c \leftarrow \bigl(\sum_u p_u\,x_u\bigr)\big/\bigl(\sum_u p_u\bigr)$
       \Comment{population-weighted centroid}
\State $P^* \leftarrow \bigl(\sum_u p_u\bigr)/N$;\enspace
       $A^* \leftarrow \bigl(\sum_u a_u\bigr)/N$
       \Comment{per-district targets}
\State $\Delta \leftarrow \mathrm{diag}\bigl(\mathrm{bbox}(\mathcal{U})\bigr)$;\enspace
       $r \leftarrow 0.001\,\Delta$
       \Comment{bounding-box diagonal; seed radius}
\vspace{4pt}
\Statex \hspace{-\algorithmicindent}\textbf{Phase 1\enspace Radial seed placement}
\For{$d = 0, \dots, N-1$}
    \State $s_d \leftarrow c + r\,\bigl(\cos(2\pi d/N),\;\sin(2\pi d/N)\bigr)$
\EndFor
\vspace{4pt}
\Statex \hspace{-\algorithmicindent}\textbf{Phase 2\enspace Capacitated first-fit assignment}
\State $\rho_i \leftarrow P^*$ for all $i$;\enspace
       $\pi(u) \leftarrow \varnothing$ for all $u$
\State Sort pairs $(u,\,i)$ by $\|x_u - s_i\|/\Delta$ ascending
       \Comment{ties: unit id, then district id}
\For{each pair $(u,\,i)$ in sorted order}
    \If{$\pi(u) = \varnothing$ \textbf{and} $\rho_i \geq p_u$}
        \State $\pi(u) \leftarrow i$;\enspace $\rho_i \leftarrow \rho_i - p_u$
    \EndIf
\EndFor
\State Assign each unassigned $u$ to $\arg\max_i\,\rho_i$
       \Comment{ties: lowest district id}
\vspace{4pt}
\Statex \hspace{-\algorithmicindent}\textbf{Phase 3\enspace Contiguity repair}
\While{$\exists\;i$ with more than one connected component}
    \State Let $C$ be the smallest component of district $i$
    \For{each $u \in C$, each district $j \neq i$ adjacent to $u$}
        \State $\mathrm{cost}(u,\,j) \leftarrow
               \dfrac{\|x_u - s_j\|}{\Delta}
               + \dfrac{|P(D_j)+p_u-P^*|}{P^*}
               + \dfrac{|A(D_j)+a_u-A^*|}{A^*}$
    \EndFor
    \State $\pi(u) \leftarrow \arg\min_j\;\mathrm{cost}(u,j)$ for each $u \in C$
           \Comment{ties: pop pen, area pen, dist, district id}
\EndWhile
\vspace{4pt}
\Statex \hspace{-\algorithmicindent}\textbf{Phase 4\enspace L\textsuperscript{1} population tightening}
\State $\delta_i \leftarrow P(D_i) - P^*$ for all $i$
\Repeat
    \For{each safe boundary move $(u,\,d_{\mathrm{dest}})$}
        \State Compute $\Delta_{L^1}(u,d_{\mathrm{dest}})$ and max-norm priority
               \Comment{priority 0 = max-reducing, 1 = $L^1$-only}
    \EndFor
    \State Accept safe top-ranked candidate; update $\pi$, $\delta$
    \If{no improving 1-opt move \textbf{and} $\max_i|\delta_i|/P^* > T$}
        \State Try chain escape of length $k=2$, then $k=3$
    \EndIf
\Until{$\max_i|\delta_i|/P^* \leq T$ \textbf{or} no improving move found}
\vspace{4pt}
\Statex \hspace{-\algorithmicindent}\textbf{Phase 5\enspace DBS hill-climb}
\Repeat
    \State Accept the safe boundary move $(u,\,d_{\mathrm{dest}})$ that
           maximally improves $\mathrm{DBS}$~(eq.~\ref{eq:dbs})
           subject to $\max_i|\delta_i|/P^* \leq$ achieved tolerance
\Until{no improving safe move exists}
\end{algorithmic}
\end{algorithm}

\subsection{Scoring}

Define per-district relative deviations
$\mathrm{pop\_dev}_i = |P(D_i) - P^*|/P^*$ and
$\mathrm{area\_dev}_i = |A(D_i) - A^*|/A^*$ and let
$\overline{\mathrm{pop\_dev}}$, $\overline{\mathrm{area\_dev}}$
denote their means over $i = 1, \ldots, N$. The DualBalance Score is
\begin{equation}
  \mathrm{DBS}
    = \frac{1}{1
       + \tfrac{1}{2}\,\overline{\mathrm{pop\_dev}}
       + \tfrac{1}{2}\,\overline{\mathrm{area\_dev}}}.
  \label{eq:dbs}
\end{equation}
The $0.5/0.5$ weighting makes the error a convex combination of the
two mean deviations: each district is judged on representing roughly
$1/N$ of the people \emph{and} roughly $1/N$ of the state's geography.
The score reaches $1.0$ for a perfectly balanced plan and approaches
$0$ as deviations grow without bound. The equal weighting is an explicit design choice, pre-committed
before any state-specific data are examined. The pre-commitment
is the critical property: once a weighting is fixed publicly,
neither the designer nor any adopting body can adjust it to favor
a particular geographic outcome. The symmetric $0.5/0.5$ choice reflects an impartiality principle:
a designer who did not know whether their state's geography would
favor population-concentrated or area-concentrated outcomes would
choose the weighting that gives neither axis an a priori advantage.
A weight of $1.0$ on population and $0.0$ on area recovers
population-only optimization; an asymmetric choice between these
poles requires a principled justification for why one representational
axis should subordinate the other within the same chamber. In the
absence of such a justification, $0.5/0.5$ is the principled default
pending democratic deliberation over the weighting itself. Secondary metrics (Polsby-Popper compactness~\cite{polsbypopper1991}
and Reock~\cite{reock1961}) are computed alongside but not optimized
against; the Phase~2 optimizer (\S\ref{sec:methods-tighten})
hill-climbs DBS, not the compactness metrics. Radial slices have
lower compactness than blob-Voronoi or hand-drawn districts by
construction; this is a deliberate trade in service of the
dual-balance objective.

DualBalance does not directly minimize~\eqref{eq:dbs}; it minimizes
population-capacitated geographic assignment cost under radial
seeding. The empirical consequences are reported in
\S\ref{sec:results}.

For partisan-fairness comparison we report the Efficiency Gap
(EG)~\cite{stephanopoulosmcghee2015}. Let $\mathrm{wasted}_P(d)$
denote the wasted votes for party $P$ in district $d$: all votes
cast for a losing candidate, plus votes above the bare majority
threshold for a winning candidate. Then
\begin{equation}
  \mathrm{EG}
    = \frac{\sum_d \mathrm{wasted}_R(d) - \sum_d \mathrm{wasted}_D(d)}
           {\sum_d \mathrm{total}(d)},
  \label{eq:eg}
\end{equation}
where positive values indicate Republican advantage and negative
values indicate Democratic advantage. EG is reported as a diagnostic.

Vote inputs for EG are drawn from the best available precinct-level
source per state, in priority order: (1)~DRA's 2016--22 composite
election (\texttt{E\_16-22\_COMP}), which blends presidential, Senate,
and gubernatorial returns across three cycles to mitigate incumbency
effects and uncontested-race artifacts~\cite{dra2020vtd}; (2)~2022
U.S. House returns (\texttt{E\_22\_CONG}) where the composite is
unavailable; (3)~2020 presidential returns (\texttt{E\_20\_PRES}) as a
fallback. The composite is available for 18 of 41 states, the 2022
House-only for 2 states, and the 2020 presidential for the remaining
21. The \texttt{votes\_source} field in each state's units file records
which source was used; rows where the fallback presidential source
applies are marked $\star$ (Table~\ref{tab:multistate-dbs}).

\subsection{\texorpdfstring{L\textsuperscript{1}}{L1} population tightening and DBS hill-climb}
\label{sec:methods-tighten}
\label{sec:methods-optimize}

The radial pipeline produces per-district pop deviation in the
5--15\% range on real census geometry, far above the strict equality
standard required for congressional districts under \textit{Wesberry
v.~Sanders}~\cite{wesberry1964} and \textit{Karcher
v.~Daggett}~\cite{karcher1983}. A deterministic two-phase local
search of boundary-unit moves closes this gap; all results reported
in this paper use this pass. Let
$\delta_i = P(D_i) - P^{*}$ denote the signed population deviation of
district $D_i$, let $T$ denote the user-supplied tolerance (default
$0.005$), and call a move \emph{safe} if the source unit is not an
articulation point of its district's induced subgraph (so that
removing it preserves contiguity).

\paragraph{Phase 1: population tightening (hybrid L\textsuperscript{1} + max-norm).}
At each step, scan every boundary unit and every neighboring district
and compute both the L$^1$ change
\[
  \Delta_{L^1}(u, d_{\mathrm{dest}}) =
    |\delta_{d_{\mathrm{src}}} - p_u| - |\delta_{d_{\mathrm{src}}}|
  + |\delta_{d_{\mathrm{dest}}} + p_u| - |\delta_{d_{\mathrm{dest}}}|
\]
and the max-norm effect: a move \emph{strictly reduces} the global
maximum iff $|\delta_{d_{\mathrm{src}}} - p_u| < \max_i |\delta_i|$,
$|\delta_{d_{\mathrm{dest}}} + p_u| < \max_i |\delta_i|$, and at least
one of $\delta_{d_{\mathrm{src}}}, \delta_{d_{\mathrm{dest}}}$ is
itself at the current maximum. Each scanned candidate is labeled with
a priority key: priority $0$ if it strictly reduces the max,
priority $1$ if it only reduces L$^1$. Candidates are ranked
lexicographically by $(\text{priority}, \Delta_{L^1})$, and the safe
top-ranked candidate is accepted.

This hybrid is essential. Pure L$^1$-greedy leaves the worst district
untouched when it sits between two other over-target districts:
draining it requires L$^1$-neutral moves that the L$^1$ criterion
rejects. Pure max-norm gets stuck symmetrically: when two districts
are tied at the max, no single move strictly reduces the global
maximum (draining one leaves the other at the same value). The
hybrid takes any move that helps either norm, with max-reducing moves
preferred so the worst district is targeted whenever possible.

When neither 1-opt criterion finds an improving move but
$\max_i |\delta_i|/P^{*} > T$, invoke a bounded \emph{chain escape}:
search for an augmenting transport on the district-adjacency graph
of the form
\[
  u_0 : D_0 \to D_1,\; u_1 : D_1 \to D_2,\; \ldots,\; u_{k-1} : D_{k-1} \to D_k,
\]
of length $k = 2$ then $k = 3$, in which each $u_j$ is a boundary unit
between $D_j$ and $D_{j+1}$. The chain is accepted if it strictly
cuts the global maximum, or if it holds the maximum flat while
strictly reducing the L$^1$ sum (which is what happens when several
districts are tied at the max: each chain drains one of them, the
global maximum stays at the same value until the last tied district
is drained, but the L$^1$ sum drops monotonically). The unit-level
search inside each district triple sorts the arc lists $B_{i \to j}$
by population value; matching pairs $(u, v)$ are found via binary
search around a target population value rather than by Cartesian
enumeration, bounding the per-triple cost by
$O(|B_{i \to j}| \log |B_{j \to k}|)$. Application is in
\emph{reverse} order ($u_{k-1}$ first, then $u_{k-2}$, etc.) so the
intermediate districts never temporarily overload during the chain.
This is the deterministic analogue of an ejection chain. Phase 1
terminates only when neither 1-opt nor bounded-chain escape finds an
improving sequence.

The L$^1$ objective is essential to the radial geometry. Seed
placement puts the most over-target and most under-target districts
on opposite sides of the population centroid, so no single
adjacent-slice swap reduces the L$^\infty$ maximum, but many such
swaps reduce the sum. A pure max-norm criterion stalls on this
geometry; the hybrid with chain escape continues to make progress
where the L$^\infty$-only criterion would halt.

\paragraph{Phase 2: DBS hill-climb.}
Once Phase 1 has converged, at each step pick the safe boundary move
that maximally improves $\mathrm{DBS}$ (equation~\ref{eq:dbs}), subject
to $\max_i |\delta_i|/P^{*}$ not exceeding the value Phase 1 reached
(equivalently, never crossing back above $T$). Phase 2 runs until no
improving safe move exists. Whereas the core DualBalance pass does not
directly minimize equation~\ref{eq:dbs}, this opt-in Phase 2 does, but
only on the post-Phase-1 plan and only within the population-compliant
feasible region.

\paragraph{Determinism and engineering.}
Both phases are pure functions of $(\mathrm{units},\,N,\,T)$; the full
pipeline including the rotation anchor is a pure function of
$(\mathrm{units},\,N,\,\theta,\,T)$ where $\theta = 0$ in all
results reported here. The only source of dependency beyond the core
pipeline's $(\mathrm{units},\,N,\,\theta)$ is the explicit tolerance $T$. The optimizer maintains a per-district articulation-point cache and
an incrementally tracked boundary-unit set, restricting each scan to
units that have a different-district neighbor (details and performance
impact at block scale in \S\ref{sec:methods-block}).

The two-phase optimizer is a concession to legal compliance; Phase 2
then directly hill-climbs the stated objective once the population
tolerance is satisfied.

\subsection{Multi-resolution refinement: VTD to Census block}
\label{sec:methods-block}

The hybrid Phase 1 + chain escape closes most of the population-balance
gap at VTD scale, but not all of it. Modern U.S. congressional plans
must clear the much tighter \textit{Karcher v.~Daggett}~\cite{karcher1983}
practical threshold of $\sim 0.05\%$ total deviation, and at VTD
granularity the residual gap is structural rather than
algorithmic. The smallest single-unit move has the size of the
smallest VTD's population. On real states the median VTD population
is on the order of $10^3$, while the \textit{Karcher} budget on a $\sim 800$k
ideal district is $\sim 400$ people. Most candidate moves overshoot
the budget on at least one endpoint, and Phase 2 starves: the
running-max constraint admits few moves, and the algorithm settles
above the \textit{Karcher} threshold with most of the residual
area-balance gain unrealized.

The fix is to refine at finer granularity. We re-run the optimizer
on Census tabulation blocks (median population $\sim 20$ rather
than $\sim 10^3$), initializing from the VTD-level \textit{Karcher}
plan rather than re-seeding from scratch. The pipeline is:

\begin{enumerate}
  \item Compute the VTD-level plan
        $\pi_{\mathrm{VTD}}: V_{\mathrm{VTD}} \to \{1, \ldots, N\}$
        via DualBalance + Phase 1 + Phase 2 at VTD scale with tolerance $T$.
  \item Load the block-level units $V_{\mathrm{block}}$ and project
        both layers to the same equal-area CRS.
  \item For each block $u \in V_{\mathrm{block}}$, find the VTD
        polygon $v(u) \in V_{\mathrm{VTD}}$ containing its
        representative point, and set
        $\pi_0(u) := \pi_{\mathrm{VTD}}(v(u))$. Blocks whose
        representative point lies outside every VTD polygon (rare,
        only on coastal/water boundary cases) fall back to the
        nearest VTD by centroid distance. The spatial join is
        deterministic given a fixed tie-breaking order on
        $\mathrm{unit\_id}$.
  \item Run Phase 1 + Phase 2 again at block scale starting from
        $\pi_0$, with the same tolerance $T$.
\end{enumerate}

Because VTDs are unions of whole blocks, $\pi_0$ inherits the
VTD-level plan's pop totals and contiguity exactly: the block-level
plan is identical to the VTD-level plan re-described at finer
granularity. Phase 1 at block scale therefore has nothing to do on
states where the VTD-level pass reached $T$. Phase 2 has substantial
freedom: block populations are two orders of magnitude smaller than
VTD populations, so the running-max envelope around $T$ admits many
candidate moves and the DBS hill-climb runs until saturation.

Because DualBalance-style seeding at block scale starts from a much worse
initial configuration than the VTD seed (the population-weighted
centroid is the same but the much higher density of degenerate
articulation-point boundaries traps the optimizer), the refinement
strategy is the path that works in practice: VTD as a coarse
solver, blocks for the precise area-balance recovery. Contiguity
checks within the optimizer use a per-district articulation-point
cache (Tarjan's algorithm~\cite{tarjan1972} on CSR adjacency arrays),
reducing each query from $\mathcal{O}(|V_d|+|E_d|)$ to
$\mathcal{O}(1)$ amortized; this is what makes the block-scale stage
tractable on commodity hardware.

